GxHash: A High-Throughput, Non-Cryptographic Hashing Algorithm Leveraging Modern CPU Capabilities

### 2023-09-22 article/article.tex
Include citation
Add description of the GxHash algorithm

\section{The GxHash Algorithm}

The \textbf{GxHash} hashing algorithm employs an high ILP construction as described in section \ref{highilp}. In practice,
this construction not only enables instruction level paralellism but also enables loop unrolling, disminishing the looping overhead,
thus maximizing computational efficiency and overall performance.\\\\
With that ILP unlocked, we still have to define f and g. In our quest for high-thoughput, it is essential for us to find computationally
efficient definitions TODO

\subsection{Pipe Width}
As for the wide-pipe construction (as seen in section \ref{widepipe}) with \( s_b \) set to the executing CPU SIMD register size.
TODO

@@

TODO

\subsection{Hashing}

TODO

\subsection{Folding}

TODO

@@

SIMD
ILP
Vectorized
Parallelized
Optimized
Throughput
Hashing

HIGH ILP CONSTRUCTION

@@

@@

@@

\section{Benchmarks}
\subsection{Setup}
\subsection{Evaluation Metrics}
\subsection{Comparison with Existing Methods}

>>>>

\clearpage
\section{The GxHash Algorithm}

The \textbf{GxHash} hashing algorithm is designed to maximize throughput by optimizing for Instruction-Level Parallelism (ILP) and making extensive use of Single Instruction, Multiple Data (SIMD) instructions. Notably, while these optimization avenues are orthogonal — focusing on parallelized and vectorized instructions respectively — they collectively harness the full potential of modern CPU architectures.\\\\
This design philosophy introduces specific constraints for the compression and finalization functions:

\begin{itemize}
\item \textbf{Utilization of Hardware Intrinsics:} To achieve the SIMD-oriented goal, arithmetic operations are tailored to be compatible with both x86 and ARM Neon intrinsics.
\item \textbf{Efficiency through Simplicity:} Minimizing the number of operations is crucial, as fewer operations typically translate to faster execution.
\item \textbf{Hash Quality Assurance:} Despite these performance optimizations, the algorithm must ensure a minimum level of hash quality to maintain low collision probabilities.
\end{itemize}

In the next sections, we'll delve into the specific operations and transformations chosen for the compression and finalization functions of the \textbf{GxHash-0} (version 0) algorithm.

\subsection{Pipe Width}

To optimize throughput, the pipe width \( s_b \) is set to match the native width of the CPU's SIMD registers.
This alignment ensures that each SIMD instruction operates on the entire pipe width, maximizing data processed per instruction.
Given that typical SIMD registers are at least 128 bits in width, the pipe width surpasses common 32-bit or 64-bit output hash sizes.
Consequently, the GxHash design aligns with the characteristics of a wide-pipe construction, as discussed in section \ref{widepipe}.
This makes GxHash more resistant to multicollision attacks.

@@

The role of the compression function is to transform a larger input (or message) into a smaller, fixed-size output. Due to this inherent reduction in size, the compression function cannot be bijective.

To delve deeper into this, consider the definition of a bijective function. A function is bijective if and only if it is both injective (one-to-one) and surjective (onto). In simpler terms, for every unique input, there is a unique output, and every possible output has a corresponding input.

Given the nature of the compression function \( f: \{0,1\}^{s_b} \times \{0,1\}^{s_b} \to \{0,1\}^{s_b} \), where the domain is much larger than the codomain (\( s_b \times s_b > s_b \)), it becomes mathematically impossible for the function to be one-to-one. There will inevitably be multiple different inputs that map to the same output, as known as collisions.\\

With the inevitable non-bijectivity, the performance requirements and the limited set of available SIMD intrinsics, the selection for the compression has to empirical, thus implying specifying a version to account for these current choice that may be improved in future versions.\\

In practice, the \textbf{GxHash-0} uses a combination of SIMD 8-bit wrapping add and state-wide circular shift of one. The addition is a simple way to mix the state with the next message block bits, while providing and arithmetic carry which helps in regards to distribution (as opposed to a XOR for instance). Adding on a 8-bit basis also helps in regards to distribution, as opposed to adding on larger bit widths. This operation alone however comes with several weaknesses:
\begin{itemize}
\item Such a simple operation is inherently weak to different kind of attacks. This can however be partially addressed with a more robust bit mixing for the finalization function. In the context of non-cryptographic hashing, which is the scope of usage of our algorithm, we think it is an acceptable compromise.
\item A major issue lies in the addition being associative (\(a \cdot b) \cdot c = a \cdot (b \cdot c \)). An associative compression function would make the hashing algorithm would be insensitive to the ordering of the input message blocks, which is something we want to avoid. To address this, we circularily shift the state bits by one.
\end{itemize}

While mathematically collisions are inevitable due to the inherent non-bijectivity, odds of collisions using the GxHash-0 compression remain statically low given the uniformity of distribution, altough those collisions are predictable due to the inherent simplicity of the function.

\begin{figure}[ht]
\begin{multicols}{2}
\begin{lstlisting}[language=Rust, style=boxed]
// For ARM 64-bit

use core::arch::aarch64::*;

pub fn compress(a: int8x16_t, b: int8x16_t)
-> int8x16_t {
    let sum: int8x16_t = vaddq_s8(a, b);
    return vextq_s8(sum, sum, 1);
}
\end{lstlisting}
\columnbreak
\begin{lstlisting}[language=Rust, style=boxed]
// For x86 64-bit

use core::arch::x86_64::*;

pub fn compress(a: __m256i, b: __m256i)
-> __m256i {
    let sum: state = _mm256_add_epi8(a, b);
    return _mm256_alignr_epi8(sum, sum, 1);
}
\end{lstlisting}
\end{multicols}
\caption{GxHash-0 Compression in Rust}
\label{fig:compression_rust_example}
\end{figure}

\subsection{Finalization}

The finalization process in the GxHash-0 algorithm is crucial to ensure the transformation of its internal state into a fixed-size, uniformly distributed hash output. This process is delineated into two primary steps: mixing the bits and folding (reducing) to the desired hash size.

\subsubsection{Mixing}

This step is responsible for ensuring the even distribution of bits in the state, thereby reducing patterns or biases that might arise from the input data or the compression process. Given the inherent simplicity of the GxHash-0 compression, it is worth for the finalization to incorporate slightly more intricate bit mixing operations, especially given it runs only once per message hashed, as opposed to the compression that occurs once for each block.\\
Leveraging SIMD capabilities can help in regards to performance and efficiency, which remains to us a primary consideration. Fortunately, both x86 and ARM architectures provide AES (Advanced Encryption Standard) intrinsics that serve as efficient tools for bit mixing. The use of an AES block cipher intrinsics ensures a robust diffusion of bits across the state at a cheap computational cost.\\
On top of that, salt is added while forming the AES block cipher keys. This not only improves the distribution but also provide a way to use random salt values per-process, protecting from eventual precomputed or replay attacks attempts.

\begin{figure}[ht]
\begin{lstlisting}[language=Rust, style=boxed]
// For x86 64-bit

use core::arch::x86_64::*;

pub fn mix(hash: state) -> state {
    // Salt (Knuth primes recommended)
    let salt = _mm256_set_epi64x(
        -4860325414534694371,
        8120763769363581797,
        -4860325414534694371,
        8120763769363581797);

    let keys = _mm256_mul_epu32(salt, hash);
    return _mm256_aesenc_epi128(hash, keys);
}
\end{lstlisting}
\caption{GxHash-0 Mixing in Rust}
\label{fig:mixing_rust_example}
\end{figure}

\subsubsection{Folding}

Once the state's bits have been thoroughly mixed, the next step is to condense or "fold" this state into a smaller, fixed-size hash output, typically 32 or 64 bits. A straightforward and effective approach taken for GxHash-0 to achieve this reduction is by summing the constituent \( X \)-bit integer parts of the mixed state. This summation serves as a reduction function, ensuring that the output hash remains within the desired size bounds while still retaining the essence of the mixed state.

While it would have been interesting to leverage SIMD once again, it turned out that, in practice, for both 128-bit and 256-bit state and for both x86 and ARM, summing the integer parts one by one is as performance if not more performant than using the various SIMD workarounds I tested.

@@

@@

Language dependent. As close to the metal as possible, avoid loop overhead.

@@

Implicit with laned construction, but worth mentionning.

@@

Load unaligned/aligned subtle difference in performance. Usually memory allocation are somehow aligned, but when using wide pipe (wide SIMD registers supported) it might not be the case that data is aligned.
Overhead for aligning data.

\section{Benchmarks}

\subsection{Quality}
\subsection{Performance}

t1ha\cite{rust-t1ha} xxhash\cite{twox-hash} HighwayHash\cite{highway-rs}

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{throughput.png}
\caption{Throughput Benchmark Results}
\label{fig:benchmark-throughput}
\end{figure}
